### 2025-09-18 paper/sec/introduction.tex
Drop numToStr from Fig 2.
Track tex-common on master.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} attempts to
address this issue through software infrastructure for ``transparent'' charts and figures. These are visual
outputs that provide interactive \emph{provenance queries}, user interactions (e.g. mousing over a visual
element) that allow readers to explore how visual features relate to data. In these approaches the
interactions are provided automatically via an underlying programming language (either a query language or
general-purpose language) which tracks dependencies as data flows through a computation.

In this paper, we extend interactive provenance queries to the natural language that accompanies the charts
and figures. We believe this is an important step because in most articles, charts and figures usually only
play a supporting role. The main claims are usually presented in natural language, language which also makes
quantitative statements and contains substantial implicit analytic content. In principle transparent
programming languages can also be used to create natural language which is bidirectly linked to data, by
computing key fragments of natural language from data. The following example, adapted from \cite{moosavi21},
shows how the language presented in this paper can be used to achieve this:

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \caption{\emph{Transparent text} links fragments of natural language to underlying data, via queries}\label{fig:scigen-example-website}
\end{figure}

>>>>

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} attempts to
address this issue through software infrastructure which transparently connections charts and figures to their
data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.} This enables visual outputs
to provide interactive \emph{provenance queries}, user interactions (e.g. mousing over a visual element) that
allow readers to explore how visual features relate to data. The key value proposition of these approaches is
that the interactions are provided automatically via an underlying programming language (query language or
general-purpose language) which tracks dependencies as data flows through a computation.

In this paper, we extend interactive provenance queries to the natural language that accompanies the charts
and figures. We believe this is an important step because in most articles, charts and figures usually only
play a supporting role. The main claims are usually presented in natural language, language which also makes
quantitative statements and contains substantial implicit analytic content. In principle transparent
programming languages can also be used to create natural language which is bidirectly linked to data, by
computing key fragments of natural language from data. The following example, adapted from \cite{zhang18} via
\cite{moosavi21}, shows how the language presented in this paper can be used to achieve this \rpnote{Need to
illustrate counterfactual/generalisability}:

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \vspace{1mm}
    \hrule
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-counterfactual-with-pointer.png}
    \caption{\emph{Transparent document} example, showing counterfactual behaviour}\label{fig:scigen-example-website}
\end{figure}
